Graph bipartitioning and you

### 2016-01-31 evolutionaryComputation/2-practical/report.tex
talking about findings

\documentclass{article}
\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{booktabs}
\usepackage{color}
\usepackage{bussproofs}
\usepackage[pdftex]{graphicx}
\DeclareGraphicsRule{*}{mps}{*}{}
\usepackage{pifont}
\usepackage{qtree}
\usepackage{emp}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{svg}
\usepackage{pgfplots}
\usepackage{pgfplotstable}

@@

Finding the paper was somewhat difficult as the original was behind a paywal.
But somoene put a copy on github so I just used that one. The algorithm 
involves a lot of cleaver book keeping:

\section{TODO}
MLS with several results
GLS -> I think the localsearch will recursively unecesrly decend into its members
because the local search is packed in the member factory. Maybe remove the
local search from the memberfactory?

\section{Results}
\subsection{Fiduccia Mathesis}
\subsubsection{Time}
\pgfplotstableread{fidRuntimes.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={fidRuntimes.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Time}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\subsubsection{Optima}
\pgfplotstableread{fidResults.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={fidResults.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Fitness (less is better)}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\subsection{Vertex swap first improvemnt}
\subsubsection{Time}
\pgfplotstableread{vertRuntimes.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={vertRuntimes.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Time}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\subsubsection{Optima}
\pgfplotstableread{vertResults.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={vertResults.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Fitness (less is better)}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{booktabs}
\usepackage{color}
\usepackage{bussproofs}
\usepackage[pdftex]{graphicx}
\DeclareGraphicsRule{*}{mps}{*}{}
\usepackage{pifont}
\usepackage{qtree}
\usepackage{emp}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{svg}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{seqsplit}

@@

Finding the paper was somewhat difficult as the original was behind a paywal.
But somoene put a copy on github so I just used that one. The algorithm 
involves a lot of cleaver book keeping. First of all the concept of gain is
introduced, this indicates how much gain there will be from moving a vertex
to the other partition. Then the vertexes are put into an array with as index
their gain. If there are more verteci at a particular index then they're
put into an linked list. To support this I introduced a container class
Cell that holds the vertex and also has an optional previous and next
element.
These cells are also put in another array based on their vertex ID. This
array is used for constant speed access when the connections of a vert their
gain needs to be updated after moving it to the other partition.
So you can just unty the ellemnt of the linked list (or if it has no
previous in my case its in the fornt of the bucket and I remove it from
there and reinsert the next element).

After knowing the basic datastructers the implementation is quite trivial
(although I had some tedious bugs such as index out of bounds errors or
next elements being wrongly updated).
It doesn't find the local optimum in one go, 4 to 5 iterations are
generally required, which is still much and much better than the first
improvment vertex swap.

\section{Own research topic}
I had several ideas for my own research to investigate, first of all I wanted
to do the really easy to implement ones, such as selection pressure and pop
sizes for GLS. But later, when waiting for results, I got curious about the
other local search I heard about in class. With several days left I decided
to try and implement the Fidducia Mathesis local search heuristic.
I think I've succeeded so my final research question is how does the 
Fidducia Mathesis algorithm compare to first
improvement vertex swap?

As can be seen in the results section the Fidducia Mathesis algorithm is
much faster than first improvement vertex swap. The main reason for this is
that a single pass will climb very steeply in valuation. It also appears
to produce better results, I think this is becuase highest gain is considered
first.

My other research topic is selection pressure, turns out that tournament is 
the best for GLS in case of fidducia mathesis, but it also takes the mot 
amount of time.

\subsection{Statistics}
The error marges in the graph are the maximum range of the results so these
are round numbers when substracted from the low or added to the high. The low
and high values are calulated with something called a confidence interval.
It alligns pretty well with the tTest results so I guess its ok.

\newpage
\section{Results}
Fidducia Mathesis dominates vertex swap first improvement in both runtime and
optima found. Combined with GLS it gets such a good score that I became
skeptical and decided to test the anwser more thoroughly.

Timewise Fidducia Mathesis dominates vertex swap first improvement. This
is because it starts with an incredibally steep hillclimb that flattens
out quite quicly. So you just need to pass over the string 4 or 5
times and then you are at the local optimum. Vertex swap first improvemt
increases linearly however. So it takes a lot more string itterations
and evaluations to get there.

The improvements found by Fidducia Mathesis are also a lot better, especially
the GLS variants. I'm not quite sure why this is the case, in fact I
added a paragraph scepticism and put a gene in there that found a solution
with 4 crossings on the original graph. Vertex swap first improvement gets
noway near to that. The lowest it found is like 280 crossings on the original
graph. I think the GLS variant doesn't find any better is because I always
start at index 0 of the string and take any improvment I can find, this
means the search space is biased. A random start could solve this issue but
then the function is no longer pure. The other bias vertex swap first
improvement has is that it either goes from left to write or from right to
left, you can't really do both. Fidducia Mathesis doesn't suffer from both
of these problem because it just picks whatever improvement has the maximum
gain (while keeping balance in mind), so its biased towards maximum gain,
which isn't a bad bias at all.

Something I found interesting is that the Fiduccia Mathesis local search
doesn't seem to respond well to ILS it just stays in the range of the lower
error marging of MLS. To verify this I cranked up the validation number to
100 (ie 100 times no better improvement found the local search end), and
increased the pertubation to 7 (because initially it looked like 1 to 5
where slowly going down).
However when you look to vertex swap first improvement you can see ILS
responding violently. In fact ILS-1 is tTest similiar to GLS:Killparents.

Selection method seems to be statistically irrelevant for the optimum in
Fidducia mathesis, although you pay a much bigger time price with both
tournament and kill parents selection. Tournament also touches the best 
result, but in other runs the other two did this to.

\newpage
\subsection{Fiduccia Mathesis}
\subsubsection{Time}
\pgfplotstableread{fidRuntimes.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={fidRuntimes.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Time}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\newpage
\subsubsection{Optima}
\pgfplotstableread{fidResults.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={fidResults.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Fitness (less is better)}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\paragraph{Sceptisicm} So these results seem a little to good to be true honestly,
especially compared to vertex swap first improvement. So I decided to catch
one of the results of GLS and throw it trough a bunch of tests to see if I 
could find anything wrong with it. The result is this one and it appears
to be OK with a fitness score of 2 (or 4 if you use the doubly connected graph).
I also checked if there was an equal amount of zeros and ones just to be sure.

\seqsplit{11100000101001111000000100110111101010000000110011100110011110101110110010111110010010010011000000101100111010011100001000111000110111110101110000010011111001110110100110110101110010101011000101000111100010111100010001100100000110001001010010101001110010111110001001010000001111011111111101101000101100000100011011101111001010010111001101101111100001011000001100100001011101000101111000101011101001001010101011011000110000110111111101101000001111000010010001111101000101000010111100101011100111100101}

There is a good chanche this is a global optimum.

\newpage
\subsection{Vertex swap first improvemnt}
\subsubsection{Time}
\pgfplotstableread{vertRuntimes.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={vertRuntimes.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Time}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};

@@

\newpage
\subsubsection{Optima}
\pgfplotstableread{vertResults.table}\dataTable
\begin{tikzpicture}
\pgfplotsset{set layers}
\begin{axis}[
	ybar stacked,
	xlabel={Algorithm},
	xticklabels from table={vertResults.table}{name},
	xtick=data,
	x tick label style={rotate=90,anchor=east},
	ylabel={Fitness (less is better)}
]
% Layers needed to get error bars on top of lower blue rectangle
\addplot[color=white]
table[on layer=axis grid, x expr=\coordindex, y=firsty]{\dataTable};
